\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-26 paper.tex

\section{Event ARGs}
\label{sec-eARG}
% The prevailing view is that ARGs are composed of events, and so we
% characterise this view
% Besides viewing ARGs as being composed of events, another
% feature of the definitions given above is that they do not
% explicitly describe
% % TODO check this! Do they talk about how trees are embedded?
% the details of how such events are represented.
Where the concrete details of an ARG data structure
(the encoding) are
% TODO more?
specified~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues.
Because it is focused on recording the details of events,
we will refer to this Griffiths encoding as an ``Event ARG'', or eARG.
In an eARG, we have two different types of node in the graph
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
taking decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
\begin{figure}
\centering
\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

@@

While the Griffiths approach of annotating recombination nodes
with a breakpoint in an eARG is a concise and elegant way of describing
realisations of the coalescent, it has some significant
issues when applied to the complexities of present-day datasets.
Because we are explicitly modelling the \emph{properties} of events,
the encoding must be extended to incorporate
different types of event that are not simple crossover recombination
or common ancestry.
Processes such as gene conversion, for example, can in principle
be included, but either require additional node types
or, for instance, multiple recombination nodes joined by a zero-length edge.
Modelling multiple recombinant offspring of the same parents
(while keeping track of their identity as siblings)
as occurs
in Wright-Fisher models (e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
or densely sampled multi-generational
data~\citep[e.g.][]{turnbull2018hundred,hayes20191000,RosFreixedes2022}
% circumlocutions isn't technically correct, but I like the roundabout
% overly verbose feel it gives.
would require more involved circumlocutions.
These and other problems can be worked around
(although, we would ideally like everyone to use the
\emph{same} workarounds if methods are to interoperate; see the Discussion),
but it is much simpler to use the gARG encoding described
in the previous section.
By recording the \emph{outcome} of events in terms of the
intervals of genetic inheritance on edges instead of recording the
properties of the events themselves, we remove the need to explicitly
model different types of event in the data structure.
More fundamentally, because gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant,
we can encode the outcome of
\emph{multiple} genetic inheritance events, and to
systematically describe the uncertainty around their
ordering (see Section~\ref{sec-precision}).

There is also a certain clarity gained by explicitly modeling nodes
in the inheritance graph as an individual's genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occured
(Appendix~\ref{sec-cell-lineages-and-args}),
or whether multiple recombinations
within a single individual should be regarded as one or multiple events are
debateable.
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
From the perspective of a concrete data structure,
ideally forming the basis of an ecosystem of interoperable
inference and analysis methods, such debates
are unproductive.

% By regarding nodes as an individual's genomes and
% explicitly annotating edges with the intervals of genetic inheritance
% between those genomes, we can capture the \emph{outcome} of a
% range of different types of event without needing to explicitly model
% the details of these events in the ARG data structure.
% More fundamentally, because gARGs
% allow us to directly describe
% genetic inheritance in multiple non-contiguous genome intervals
% between an ancestor and a descendant,
% we can encode the outcome of
% \emph{multiple} genetic inheritance events, and to
% systematically describe the uncertainty around their
% ordering (see Section~\ref{sec-precision}).

% While the additional generality provided by the gARG representation
% is useful, it is in reality a small change that does not
% lead to significantly different time or space complexity in computational
% tasks.
% The vitally important and fundamental distinction between
% a gARG and an eARG is that gARGs
% allow us to directly describe
% genetic inheritance in multiple non-contiguous genome intervals
% between an ancestor and a descendant, which cannot be done
% by specifying recombination events and breakpoints.
% % FIXME this sentence is problematic. We should recast to talk about
% % the outcomes of events, not talk about simulateousness (which will
% % confuse people
% Thus, we have the means to encode the \emph{outcome} of
% multiple genetic inheritance events, and to
% systematically describe the uncertainty around their
% ordering (see Section~\ref{sec-precision}).
% As we describe in the following sections, freeing ourselves from the requirement
% of recording specific events and
% the additional generality of the gARG encoding
% has significant advantages. It also raises questions about what our goals
% for inferring ARGs should be, and how we evaluate the success
% of such inferences.

% A common feature of these definitions is that they are expressed
% in terms of evolutionary \emph{events}, and so for the purposes of
% discussion we can label
% this prevailing interpretation as an ``event ARG'', or eARG.
% (Many definitions also emphasise
% the necessity of exhaustively capturing \emph{all} events, which we
% examine in later sections.)
% Another common feature of these definitions is that they do not explicitly
% discuss the mechanism by which local trees vary along the genome.
% The topology of an eARG defines the common ancestor and
% recombination events that occurred,
% but the graph structure alone is not sufficient to define the local trees,
% or equivalently, how ``ancestral material'' flows along
% the edges of the ARG (see Section~\ref{sec-eARG-to-gARG}).

% TODO reference forward to the cell ARG appendix. Ultimately
% every event corresponds to a particular genome in a particular
% individual or cell.
% Whether we regard nodes as events or genomes is something of a
% philosophical question, and in most senses the two are equivalent
% (but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
% From a concrete data-structure encoding perspective, the only real
% difference between an eARG and a gARG is that in an eARG we store recombination
% \emph{breakpoints} at nodes of out-degree two (recombination nodes)
% whereas in a gARG we store \emph{inheritance intervals} on edges.
% A key advantage of associating inheritance intervals with
% edges in a gARG is that it allows us to describe genetic inheritance events other
% than simple crossovers, such as multiple recombinations and
% gene conversions. By explicitly stating the exact regions of genome
% that are transmitted along an edge, we generalise the representation
% to encompass any process that leads to the differential inheritance of
% genetic material along the genome.

% While the additional generality provided by the gARG representation
% is useful, it is in reality a small change that does not
% lead to significantly different time or space complexity in computational
% tasks.
% The vitally important and fundamental distinction between
% a gARG and an eARG is that gARGs
% allow us to directly describe
% genetic inheritance in multiple non-contiguous genome intervals
% between an ancestor and a descendant, which cannot be done
% by specifying recombination events and breakpoints.
% % FIXME this sentence is problematic. We should recast to talk about
% % the outcomes of events, not talk about simulateousness (which will
% % confuse people
% Thus, we have the means to encode the \emph{outcome} of
% multiple genetic inheritance events, and to
% systematically describe the uncertainty around their
% ordering (see Section~\ref{sec-precision}).
% As we describe in the following sections, freeing ourselves from the requirement
% of recording specific events and
% the additional generality of the gARG encoding
% has significant advantages. It also raises questions about what our goals
% for inferring ARGs should be, and how we evaluate the success
% of such inferences.

>>>>

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where such concrete details of an ARG data structure
(the encoding) are
% TODO more?
mathematically defined~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and the majority of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[(TODO exhaustive list)][]{rasmussen2014genome,ignatieva2021kwarg}.
In this Griffiths encoding, we have two different types of node in the graph
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``Event ARG'', or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

\begin{figure}
\centering
\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

@@

% eARGs are representationally for limited interchange
While the Griffiths approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it has some significant issues when applied to the
complexities of present-day datasets. The eARG encoding explicitly models only
two different types of event and thus anything that is not a single crossover
recombination or common ancestor event, must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. From the perspective of practical interchange of data between
inference methods and downstream applications, both
workarounds are problematic, and the gARG encoding described in the previous
section offers a much simpler solution.
Modern datasets contain many more complexities which the eARG encoding
is ill-suited to represent (see Discussion).

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred
(Appendix~\ref{sec-cell-lineages-and-args}),
or whether multiple recombinations
within a single individual should be regarded as one or multiple events are
debatable.
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
From the perspective of a concrete data structure,
ideally forming the basis of an ecosystem of interoperable
inference and analysis methods, such debates
are unproductive.
